Model based reinforcement learning at FERMI FEL

### 2020-11-17 main.tex
txs auto checkin

\begin{abstract}
In this paper we discuss a model based reinforcement learning approach in comparison to a model free reinforcement learning approach applied at the FERMI FEL system. Both algorithms and approaches are new in this context and the main purpose of this paper is to show that reinforcement learning can be applied on an operational level in a feasible training time on real accelerator physics problems. In terms of sample-complexity the model-based approach is faster, while the finale performance of the model free method is superior - the so called asymptotic performance. The model-based algorithm is done in a Dyna-style using an uncertainty aware model and the model-free algorithm is based on tailored deep Q-learning using some tricks to increase the sample efficiency
\end{abstract}
%\keywords{Suggested keywords}%Use showkeys class option if keyword
                              %display desired
\maketitle

%\tableofcontents
\section{Introduction and Motivation}
In particle accelerators one main goal is to provide stable and reproducible performance. In order to achieve this, a number of control problems have to be considered simultaneously. Especially if there is no way to model the physics, one might use optimisation techniques as e.g. derivative free optimisers (DFOs) or model-based optimisations as Gaussian processes. Another promising way is to apply reinforcement learning which has several advantages over optimisation methods:
\begin{itemize}
    \item Covers a larger class of problems.
    \item Solves the problem not always completely from the beginning as an DFO.
    \item Ability to use existing data.
    \item The underlying structure of the problem might be deduced.
\end{itemize}
RL optimises a sequence of decisions. On critical aspect using RL is the number of iteration to train a controller.
In this paper, we present the study carried out to solve the maximisation-problem of the radiation intensity generated by a seeded Free-Electron Laser (FEL) on the Free Electron laser Radiation for Multidisciplinary Investigations (FERMI) at Elettra Sincrotrone Trieste. Three algorithms were applied successfully showing different advantages, which will e discussed in the following.

@@

\section{Deep Reinforcement learning}
Assume states $s$ the set of all states $\bs$ and actions $\ba$ all actions $\actions$ and rewards $\reward$ and the set of all rewards $\mathcal{R}$ an initial state distribution $\initstate$.
Goal of reinforcement learning is to find a $\policy\mapsto a$, which is the solution of:
\begin{align}
\max J (\policy)  =  \mathbb E_{\traj \sim p_{\policy}}\left[\sum_{t=0}^{\horizon}\discount^t r(\bs_t,\ba_t)\right],
\end{align}
where $p_\policy = \initstate(\bs_0)\prod_{t_0}^\horizon\policy(\ba_t,\bs_t)\transitions(\bs_{t+1}|\bs_t,\ba_t)$ is the distribution of all trajectories $\traj := (\bs_0, \ba_0, \bs_1, a_1,\dots s_\horizon,a_\horizon)$ drawn by $\policy$ with a horizon $\horizon$.
In the modern field of RL one distinguishes if the policy $\policy(\ba)\approx \policy_\phi(\bs)$ or the state-action value function $Q(\bs,\ba)$ is approximated using a high capacity function approximator, as e.g. a deep neural network. In the first case one speaks about policy gradient methods in the latter about approximate dynamic programming, which we now discuss.
\subsection{Approximate dynamic programming}
The state-value function $Q(\bs,\ba)$ is expressed as $Q_\theta(\bs,\ba)$, where $\theta$ denotes the parameters of the function approximator. By satisfying the Bellmann-optimality operator the optimal $Q^*(\bs,\ba)$ can be found:
\begin{align}
    \min_\theta \left(\vec Q_\theta - \bellman^*\vec Q_\theta\right)^2.\label{eq:minimize_bellmann_optimality}
\end{align}
This operator has a unique fixed point but is non-linear, due to the $\max$ - operator:
\begin{align}
 \bellman^*\vec Q_\theta(\bs_t,\ba_t) := r(\bs_t,\ba_t)+\gamma\max_\ba\left( Q_\theta(\bs_{t+1},\ba_t) - Q_\theta(\bs_t,\ba_t)\right).
\end{align}
The form of this equation can cause overestimation and other complications, when using a function approximator. Several methods exist which try to mitigate the problems \cite{}.
One way to reduce the effect is to take a simple analytical form of the $Q$-function.
\subsection{NAF}
If a specific quadratic form of the $Q$ function is assumed:
\begin{align}
     Q_\theta(\bs,\ba) = -\frac{1}{2}(\ba-\mu_\theta(\bs))P_\theta(\ba-\mu_\theta(\bs))^T+V_\theta(s).
\end{align}
 The policy is given by:
\begin{align}
    \policy_\theta(\bs)=(\delta(\ba)-\arg\!\max_\ba Q_\theta(\bs,\ba)).
\end{align}
One modification, which is used in these tests is a twin network (weights denoted by $\tilde\theta$), where only one is used to obtain the policy, while the other is used for the update rule to avoid over-estimation.
The maximum is given analytically as $\max_a Q(s,a) = V(s)$, hence from \cref{eq:minimize_bellmann_optimality}:
\begin{align}
	\min_\theta\left( (\reward(\bs_t,\ba_t)+\gamma \min_{\hat\theta,\tilde \theta} V_{\theta\prime}(s_{t+1}) - (1+\gamma) Q_\theta(s_t,a_t)\right)^2
\end{align}
$\hat\theta$ are the weights of a previous iteration or the initialized weights. To stabilize the network training a small artificial noise is added to the actions in the update. This algorithm has an extremely good sample-efficiency for suitable problems as can be found in accelerators and is used as the baseline for the considered control problem.
%\subsection{New NAF network architecture}

@@

The anchoring ensemble methods usually yields good results already with a small number of different networks. The idea behind it, is to mimic the posterior probability of the dynamics model to capture first of all its own uncertainty due to the lack of data. Empirical results show that three to five models were sufficient to see clear advantages over a single network approach and the main goal is not to calculate the exact posterior of the model. A small two-layer network with around 15-25 nodes and the $tanh$ activation were used. The last layer was linear and the inputs were normalized to the interval [-1,1]. The prior is controlled by the initialization of the weights, which were normalized by the number of nodes in each layer. The noise level is also respected during the initialization. For the moment only homeoscetastic errors are used. Several methods to set the training method were implemented. Early stopping, a standard techniques showed good results, mainly to the fact, that the number of training steps is increasing with more data. The advantage is that the uncertainty at regions without data are shrink in this way. A fixed loss value threshold was also tested, as also used in the original anchore-ensemble technique. In the experiments a combination of both was taken.\\

>>>>

\begin{abstract}
In this paper we discuss a model based reinforcement learning approach in comparison to a model free reinforcement learning approach applied at the FERMI FEL system. Both algorithms and approaches are new in this context and the main purpose of this paper is to show that reinforcement learning can be applied on an operational level in a feasible training time on real accelerator physics problems. In terms of sample-complexity the model-based approach is faster, while the finale performance of the model free method is superior - the so called asymptotic performance. The model-based algorithm is done in a Dyna-style using an uncertainty aware model and the model-free algorithm is based on tailored deep Q-learning using some tricks to increase the sample efficiency
\end{abstract}
%\keywords{Suggested keywords}%Use showkeys class option if keyword
                              %display desired
\maketitle
\begin{itemize}
	\item Noise benchmarks with pendulum
	\item Benchmark NAF2 with pendulum

\end{itemize}
%\tableofcontents
\section{Introduction and Motivation}
In particle accelerators one main goal is to provide stable and reproducible performance. In order to achieve this, a number of control problems have to be considered simultaneously. Especially if there is no way to model the physics, one might use optimisation techniques as e.g. derivative free optimisers (DFOs) or model-based optimisations as Gaussian processes. Another promising way is to apply reinforcement learning which has several advantages over optimisation methods:
\begin{itemize}
    \item Covers a larger class of problems.
    \item Solves the problem not always completely from the beginning as an DFO.
    \item Ability to use existing data.
    \item The underlying structure of the problem might be deduced.
\end{itemize}
RL optimises a sequence of decisions. On critical aspect using RL is the number of iteration to train a controller.
In this paper, we present the study carried out to solve the maximisation-problem of the radiation intensity generated by a seeded Free-Electron Laser (FEL) on the Free Electron laser Radiation for Multidisciplinary Investigations (FERMI) at Elettra Sincrotrone Trieste. Three algorithms were applied successfully showing different advantages, which will e discussed in the following.

@@

\section{Deep Reinforcement learning}
Assume states $s$ the set of all states $\bs$ and actions $\ba$ all actions $\actions$ and rewards $\reward$ and the set of all rewards $\mathcal{R}$ an initial state distribution $\initstate$.
Goal of reinforcement learning is to find a $\policy\mapsto a$, which is the solution of:
\begin{align}
\max J (\policy)  =  \mathbb E_{\traj \sim p_{\policy}}\left[\sum_{t=0}^{\horizon}\discount^t r(\bs_t,\ba_t)\right],
\end{align}
where $p_\policy = \initstate(\bs_0)\prod_{t_0}^\horizon\policy(\ba_t,\bs_t)\transitions(\bs_{t+1}|\bs_t,\ba_t)$ is the distribution of all trajectories $\traj := (\bs_0, \ba_0, \bs_1, a_1,\dots s_\horizon,a_\horizon)$ drawn by $\policy$ with a horizon $\horizon$.
In the modern field of RL one distinguishes if the policy $\policy(\ba)\approx \policy_\phi(\bs)$ or the state-action value function $Q(\bs,\ba)$ is approximated using a high capacity function approximator, as e.g. a deep neural network. In the first case one speaks about policy gradient methods in the latter about approximate dynamic programming, which we now discuss.
\subsection{Approximate dynamic programming}
The state-value function $Q(\bs,\ba)$ is expressed as $Q_\theta(\bs,\ba)$, where $\theta$ denotes the parameters of the function approximator. By satisfying the Bellmann-optimality operator the optimal $Q^*(\bs,\ba)$ can be found:
\begin{align}
    \min_\theta \left(\vec Q_\theta - \bellman^*\vec Q_\theta\right)^2.\label{eq:minimize_bellmann_optimality}
\end{align}
This operator has a unique fixed point but is non-linear, due to the $\max$ - operator:
\begin{align}
 \bellman^*\vec Q_\theta(\bs_t,\ba_t) := r(\bs_t,\ba_t)+\gamma\max_\ba\left( Q_\theta(\bs_{t+1},\ba) - Q_\theta(\bs_t,\ba_t)\right).
\end{align}
The form of this equation can cause overestimation and other complications, when using a function approximator. Several methods exist which try to mitigate the problems \cite{}.
One way to reduce the effect is to take a simple analytical form of the $Q$-function.
\subsection{NAF}
If a specific quadratic form of the $Q$ function is assumed:
\begin{align}
     Q_\theta(\bs,\ba) = -\frac{1}{2}(\ba-\mu_\theta(\bs))P_\theta(\ba-\mu_\theta(\bs))^T+V_\theta(s).
\end{align}
 The policy is given by:
\begin{align}
    \policy_\theta(\bs)=(\delta(\ba)-\arg\!\max_\ba Q_\theta(\bs,\ba)).
\end{align}
One modification, which is used in these tests is a twin network (weights denoted by $\tilde\theta$), where only one is used to obtain the policy, while the other is used for the update rule to avoid over-estimation.
The maximum is given analytically as $\max_a Q(s,a) = V(s)$, hence from \cref{eq:minimize_bellmann_optimality}:
\begin{align}
	\min_\theta\left( (\reward(\bs_t,\ba_t)+\gamma \min_{1,2} V_{\theta^i_\text{targ}}(\bs_{t+1}) - (1+\gamma) Q_\theta(\bs_t,\ba_t)\right)^2
\end{align}
$\theta_\text{targ}$ are the weights of a target network, which is softly updated. To stabilize the network training a small artificial noise is added to the actions in the update. This algorithm has an extremely good sample-efficiency for suitable problems as can be found in accelerators and is used as the baseline for the considered control problem.
%\subsection{New NAF network architecture}

@@

The anchoring ensemble methods usually yields good results already with a small number of different networks. The idea behind it, is to mimic the posterior probability of the dynamics model to capture first of all its own uncertainty due to the lack of data. Empirical results show that three to five models were sufficient to see clear advantages over a single network approach and the main goal is not to calculate the exact posterior of the model. A small two-layer network with around 15-25 nodes and the $tanh$ activation were used. The last layer was linear and the inputs were normalized to the interval [-1,1]. The prior is controlled by the initialization of the weights, which were normalized by the number of nodes in each layer. The noise level is also respected during the initialization. For the moment only homeoscetastic errors are used. Several methods to set the training method were implemented. Early stopping, a standard techniques showed good results, mainly to the fact, that the number of training steps is increasing with more data. The advantage is that the uncertainty at regions without data shrink in this way leading to a better training performance. A fixed loss value threshold was also tested, as also used in the original anchore-ensemble technique. In the experiments a combination of both was taken.\\
